Constraining the High Density Nuclear Symmetry Energy with Pions

### 2020-06-10 chapters/ch_data_analysis.tex
Backup

The S$\pi$RITROOT software is modular tasked based code based on the FAIRROOT package written in C++ \cite{fairroot}. The main tasks in the S$\pi$RITROOT software reconstruction are:
\begin{itemize}
  \item Decoder task
  \item Pulse Shape Algorithm (PSA Task)
  \item Helix Track Finding Algorithm
  \item Clustering Algorithm
  \item Track Fitting (GENFIT package)
  \item Vertex Fitting (RAVE package)
\end{itemize}

\subsection{Decoder task}
The decoder task reads in the raw the binary data output from the GET electronics and separates the data from the headers using the GETDecoder CITE HERE. The raw data is mapped from the cobo layout on the TPC to each individual channel and the information is stored in the STPad class. This is a container class which stores the row and layer information, and the raw ADC time bucket spectrum. All of the pads combined form the STRawEvent class which represents event information.

Three basic calibrations are performed at the raw data level within this class and will be described in more detail later. The gain calibration calibrates the electronics described in Section\ref{sec:elecCalib} and the anode gain calibration discussed in Section~\ref{sec:anodeCalib}. The gating grid noise subtraction and time bucket cut which is described in Section~\ref{sec:ggnoisesub}.
%gain calibration 
%ggnoise subtraction
%GG time bucket cut

\subsection{Pulse Shape Algorithm (PSA) Task}
\label{sec:psa}
%Yoffset match the TPC-Vertex_y with BDC_Y
%YPedestalOffset what is it for?
%cobo gain matching
An algorithm loops over the raw time bucket spectrum to find the unique pules in each pad. It is very likely there are several pulses in a pad coming from multiple tracks passing under the same pad separated only in arrival time. Using an expected pulse shape the Pulse Shape Algorithm (PSA) task finds the first peak in the spectrum, fits the peak with the standard pulse shape. If the fit is successful the fitted pulse is subtracted from the raw ADC spectrum and the height of the pulse $Q$, which is related to the total charge, and the timing information $t_o$ is saved into a new container class which represent a ``hit", called STHit. The algorithm continues through the subtracted spectrum finding all the pules and assigning the fitted values to new hits until there are no more pules in a spectrum. Figure~\ref{fig:psaTask} shows the PSA task working on one particular pad in real experimental data. The shaded histogram in every sub-figure represents the ADC raw spectrum of the pad. In Fig.~\ref{fig:psa0} we see the PSA has identified the first pulse and fitted it with the standard pulse, the result of the fit is shown in the red line. The fitted red line is then subtracted from the ADC spectrum which is shown in the next Fig.~\ref{fig:psa1}. Here the second pulse is found and fitted again with the same procedure repeating were the red line will be subtracted from the raw ADC spectrum. Here the black line shows the sum of all the fitted pules. This continues for every pules in the spectrum until all pules are found or until it reaches the last bin in the spectrum. Figure~\ref{fig:psa4} shows the final result of all 4 fitted pules and their sum which is given in the black line. The agreement between the total summed line and the raw ADC spectrum is very good.

After the PSA processes each pad in the raw event, an entire hit cloud of the event is produced. The x and z-coordinates come from the center of the pad which is already stored in the map of STPad. Since the STHit class inherits from the STPad class this information is also carried over. The y-coordinate is calculated using the known drift velocity $v$ and time of the signal $t_o$ as $y = v\cdot t_0$, $t_0$. The hit class forms the fundamental structure from which all tracking is performed since it contains the x,y,z spacial coordinates, and charge information $Q$.

\begin{figure}[!htb]
     \centering
     \begin{subfigure}[b]{0.24\textwidth}
         \centering
         \includegraphics[width=\textwidth]{psa0.png}
         \caption{Fist step of the PSA}
         \label{fig:psa0}
     \end{subfigure}
     \hfill
     \centering
     \begin{subfigure}[b]{0.24\textwidth}
         \centering
         \includegraphics[width=\textwidth]{psa1.png}
         \caption{Second iteration of the PSA}
         \label{fig:psa1}
     \end{subfigure}
     \hfill
      \centering
     \begin{subfigure}[b]{0.24\textwidth}
         \centering
         \includegraphics[width=\textwidth]{psa2.png}
         \caption{Third iteration of the PSA}
         \label{fig:psa2}
     \end{subfigure}
     \hfill
      \centering
     \begin{subfigure}[b]{0.24\textwidth}
         \centering
         \includegraphics[width=\textwidth]{psa3.png}
         \caption{Fourth iteration of the PSA}
         \label{fig:psa3}
     \end{subfigure}
     \hfill
     \centering
     \begin{subfigure}[b]{.5\textwidth}
         \centering
         \includegraphics[width=\textwidth]{psa4.png}
         \caption{Final output of the PSA}
         \label{fig:psa4}
     \end{subfigure}
     \hfill

@@

\subsection{Helix Tracking}
%Cluster cut
%ellipsoid cut

 The function of the  helix track finding algorithm is to associate the collection of hits belonging to one track out of all the entire hit cloud produced by the PSA. This is performed by a Reimann track finding algorithm where the cartesian space is mapped onto a Reimann sphere. This is a particular transformation in which unique helices in the Cartesian space form unique circles on the Riemann sphere. This is particularly useful since hits which correspond to a track form a helix the Cartesian space of the TPC coordinates. It is quite difficult to search the hit cloud space fitting a general helix with different combinations of hits. It is much easier to search for collection of hits by fitting a plane to the unique great circles formed on the Riemann sphere. From here the hits which belong to unique circles are set into a new helix track container called the STHelixTrack.

\paragraph{Definition of clustering}
The hits within a helix track are dense and the position resolution of an individual hit is not precise. As mentioned in Section~\ref{sec:padResponse}, due to the PRF, the collection of hits perpendicular to the track gives the TPC sub millimeter position resolution. In the helix track finding task, the hits are further reduced into clusters. A brief description of the method of clustering is illustrated in Figure \ref{fig:topview}. It is impractical to cluster in both the x and z-axis at the same time and instead we only cluster in one direction. For example, the three clusters at the bottom of Figure \ref{fig:topview} are clustered along the x-axis and the upper three are along the z-axis. The bolded pads highlight an example from each set. The direction perpendicular to the track is the direction of the highest position resolution for a track. Since we decide to cluster only in one direction, there is a natural inflection point in a track at which the direction of clustering switches. This  depends on the crossing angle $\theta$, which is defined as the angle the tangent makes with respects to the x-axis. In this definition $\theta = \ang{90}$ when the track is going along the z-axis, and $\theta = \ang{0}$ for a track going along the x-axis. In the case that the crossing angle is $\ang{45} < \theta \leq \ang{90} $ the we cluster along the x-direction. For $\ang{0} < \theta \leq \ang{45}$ we cluster along the z-axis.

%A cluster's position is the average, charge-weighted position of the hits within a cluster and the total charge of the cluster is the sum of all the hits charges.

 %A tracks average position is estimated by the cluster's average position. The clusters are then fitted in the GENFIT track fitting package \cite{genfit}, giving the final momentum of the track. A final vertex of the event is fitted from all tracks using the package RAVE \cite{rave}. 
 The position along the clustering direction, $\overline{X}$, is calculated as the charge-weighted average, $\overline{X} = \sum_i q_ix_i$, where $q_i$ and $x_i$ are the charge and position of the i-th hit. The other direction is set to the center of the pad. For example, if we are clustering hits along the x-axis, the z-position is set to the center of the pad in the z-direction and vice versa. Clustering in this way gives us better position resolution for calculating the position of each cluster. You can imagine if we calculated the clusters only along the x-axis for tracks with $\theta \approx 0^{\circ}$ the x-position is not well defined.

\begin{figure}[H]
\includegraphics[scale=.5]{top_view_helix_ext.pdf}
\caption{Cartoon graphic of a top down view of a fit to a track passing through several pads. The bolded pads and the charges $q_i$ represent the hits belonging to that pad and the clusters of the track representing the average position of the track. The three clusters at the bottom are clustered in the x-direction and for the upper three clustered in the z-direction. The estimate of the position of the avalanche is given by the track fit and the position from the center to each pad to the $\bar{x}$ position is given as $\lambda_i$.}
\label{fig:topview}
\end{figure}

\subsection{Correction Task}

\subsection{Space charge}

\subsection{Pad response function cut}

\subsection{Momentum and track reconstruction (GENFIT)}

\subsection{Vertex tracking (RAVE)}
\label{sec:vertex
}

\section{Calibrations and Corrections}

\subsection{Gating grid noise subtraction}
\label{sec:ggnoisesub}
Opening the gating grid essentially short adjacent wires together allowing them to reach equilibrium as fast as possible. In practice the impedance of both sides was not entirely matched properly and this caused an oscillating current to bounce back and forth in an under-dampened manner. This caused residual induced signals early in the time bucket spectrum which created and extra source of noise in the data. The ground grid shielded some of the noise, but the path to ground was on two ends and was not sufficiency for this frequency. Figure~\ref{fig:ggNoiseSub} shows in the upper panel the gating grid noise ADC time bucket spectrum for 2000 events in a given pad. The signal is stable within a pad and the mean value --shown as the red line-- can be calculated after averaging over several thousands events. The raw gating grid noise lasts for 100 tbs extending into the real data with a decreasing amplitude.

@@

The gating grid itself was not always stable an only the ${}^{132}$Sn and ${}^{124}$Sn beams were relatively stable. Several times during the ${}^{108}$Sn and ${}^{112}$Sn beams the gating grid broke. After several data runs were taken, we took what is called gating grid noise runs to get the profile of the gating grid noise throughout the experiment. To achieve this the anode wires would be turned off and the trigger was allowed to otherwise operate normally, firing the gating grid to open as usual. Since the anode wires were all turned off no signals would be seen except those coming from the gating grid noise. Several thousand triggers of the gating grid were taken to get a good noise profile.

Once the mean value response in each pad is calculated, we can use the gating grid noise profile runs to correct the real data and remove the noise of the gating grid. To show the best case scenario of this technique, the bottom panel of Fig.~\ref{fig:ggNoiseSub} shows the gating grid noise self correcting itself. We can see the pedestal is already subtracted naturally when using this technique, and the subtracted ADC spectra is centered around 0. Residual gating grid noise that was not canceled by the subtraction remains but is significantly reduced to a much smaller time bucket region. This happens in regions where the gating grid noise is changing rapidly, where the ADC value is dependent upon small jitters in the timing of the signal. All time buckets below 30 tb were cut away from the spectrum to get rid of the small region where the gating grid is still opening and to remove the residual noise.

@@

\subsection{Cocktail calibration}

\begin{table*}\centering
\ra{1.3}
\begin{tabular}{@{}rrrrrrr@{}}\toprule
& \multicolumn{3}{c}{$100 MeV Target$}\\
\cmidrule{2-4}
Particle &\phantom{abc} & Measured & Corrected & \% Difference & \% Difference\\
\midrule
p   & 882.8 & 929.5 & 877.3   &  3.7  & -1.0  \\
d   & 817.1 & 831.15 & 797.94 &  1.7  & -2.3\\
\bottomrule
\end{tabular}
\caption{Summary of expected cocktail. }
\label{tb:cocktail100tar}
\end{table*}

\begin{table*}\centering
\ra{1.3}
\begin{tabular}{@{}rrrrrrr@{}}\toprule
& \multicolumn{3}{c}{$100 MeV$}\\
\cmidrule{2-4}
Particle & Expected & Measured & Corrected & \% Difference & \% Difference\\
\midrule
p   & 882.8 & 903.5 & 889   &  2.0   & -1.6  \\
d   & 817.1 & 898.5 & 874.5 &  2.1   & -2.7\\
\bottomrule
\end{tabular}
\caption{Summary of expected cocktail. }
\label{tb:cocktail100}
\end{table*}

@@

\bottomrule
\end{tabular}
\caption{Summary of expected cocktail. }
\label{tb:cocktail300}
\end{table*}

A cocktail of several light charged particles were produced (p,d,t,${}^{3}$He,${}^{4}$He), and measured in the TPC to provide a momentum calibration of the TPC. The magnetic and slit settings of the dipoles in the BIGRIPS spectrometer was set so that the measured momentum resolution of the beam was $\delta p/p$ < 1\%. Two magnetic rigidity settings were studied, with an empty target. A thick Aluminum target was used to provide a slightly lower energy point for part of the lower rigidity setting, effectively creating three calibration points over several particle species. The production of certain particles (t,${}^{3}$He), produced too few counts to make a good measurement, and in the high momentum rigidity setting protons could not propagate down the line and there were no counts.

Since the expected momentum resolution resulting from the spectrometer was less than 1\%, the observed momentum resolution measured by the TPC is a good representation of the combined momentum resolution of the software and TPC (intrinsic detection) system. The momentum resolution depends on several factors such as the particle's angle, momentum, charge, track multiplicity, etc. This calibration beam represents and ideal situation where the track was parallel to the pad plane and only one particle was measured at one time. The energy loss resolution can also be directly measured since each energy setting represents a monochromatic source of each particle species, which has a well defined energy loss distribution. An average momentum resolution of $dp/p = 2\%$ and the energy loss resolution of $d\langle dE/dx \rangle/\langle dE/dx \rangle = 5\%$ was measured for particles ranging from $H$ to ${}^{4}$He, over the range of momenta measured in the calibration beam as summarized in Table~\ref{tb:momresolution}.

Since the magnetic dipole setting of the BIGRIPS spectrometer defined the energy of each particle type, we can calculate the expected momenta of each particle species measured. Small corrections were propagated using LISE++ software which can calculate the energy loss through several materials in the beam line. These corrections resulted in a small change in the momenta. The measured momenta of the calibration beam differed significantly from the expected values as seen in Tables~\cref{tb:cocktail100tar,tb:cocktail100,tb:cocktail300}. This effect is attributed to inhomogenatities in the magnetic field which introduces electron drift velocity in the direction of $\vec{E}\times\vec{B}$ direction as seen from Eq.\ref{eq:elecdrift}. The $\vec{E}\times\vec{B}$ drift velocity causes the electron trajectories to shift toward the +x-axis in the TPC coordinates causing particles of positive charge (going in the -x-axis) to have a higher measured momenta than in reality. The disagreement in measured and expected momenta is upwards of ~5\% difference in the higher momentum calibration settings. The details of the correcting for the $\vec{E}\times\vec{B}$ effect will be discussed in the later Section~\ref{sec:spacecharge} in a more general way which also includes correction for the space charge, which this cocktail beam does not have. The same correction technique was applied here in the special case of zero space charge, which is the special case of only having $\vec{E}\times\vec{B}$ components.

The values under the corrected column seen in Tables~\cref{tb:cocktail100tar,tb:cocktail100,tb:cocktail300}, represent the data corrected for the $\vec{E}\times\vec{B}$ effect. A significant improvement is seen in the high momentum setting going from around ~5\% disagreement to within ~1\% agreement in the corrected data. For the lower momentum settings (Tables~\cref{tb:cocktail100tar,tb:cocktail100}), protons see a slight improvement of about ~1\% where as the deuterons are over corrected in both settings. The level of agreement of the all corrected values is still within the estimated momentum resolution of the TPC.

\begin{table*}\centering
\ra{1.3}
\begin{tabular}{@{}rr@{}}\toprule
Momentum Resolution \% & <dE/dx> Resolution \% \\
\midrule
1.6  & 4.6\\
\bottomrule
\end{tabular}
\caption{Summary of expected cocktail. }
\label{tb:momresolution}
\end{table*}

@@

\subsection{Electronics calibration}
\label{sec:elecCalib}
The ADC channel number of the electronics was calibrated by measuring the response of each channel to an input pulse supplied by a pulse generator. The pulse was distributed to all the electronics channels by pulsing the ground plane for a range of input voltages. This distributed the pulse evenly across the entire pad plane. The input voltage is plotted as a function of the measured ADC channel in Fig.~\ref{fig:gaincalib} for every channel. The small variation in each channel can be seen as the wide band around each measurement point. A linear fit is performed to get the best fit line which provides a reference line which each channel is calibrated to. The right panel shows the resulting distribution of channels after calibration. This is a relative calibration technique with the intent to calibrate the varying gains in each channel relative to one another.

\begin{figure}[H]
\includegraphics[width=\linewidth]{gaincalib.png}
\caption{Calibration of electronics}
\label{fig:gaincalib}
\end{figure}

\subsection{Anode gain calibration}
\label{sec:anodeCalib}
As mentioned earlier, the anode wires were separated into 14 independently biased sections. The high voltage of sections 12 and 14 were reduced during the experiment due to high currents being observed on the wires. The wire section voltages were lowered once and adjusted once again. Out of all the runs used in the analysis in this thesis \ref{tb:runList}, the anode sections 12 and 14 were lowered to \SI{1085}{\volt} for runs 2272-2371 and set to \SI{1214}{\volt} for all the other runs. By lowering the voltage on these anode wires, the gas gain is lowered as compared with all the other anode wire plane sections which operate at \SI{1460}{\volt}. To account for the drop in gain, in the software we increase the gain of the pads which lie above these anode wires to match the gain in the other channels. To do so, we compare the energy loss in the high gain sections to the low gain sections to get the relative calibration factor.

>>>>

The S$\pi$RITROOT software is modular code based on the FAIRROOT package written in C++ \cite{fairroot}. The code is separated into modules which can be turned on or off depending on the users needs. The base classes and frame work is provided by the FAIRROOT package, while the specifics of each detector need to be added into the base classes. In the following an outline of the main tasks in the S$\pi$RITROOT software will be briefly discussed with the details of each task discussed later. The discussion of the software is really a practical, base understanding, appropriate for this thesis. For more details about the tracking algorithm the reader is refered to \cite{spiritroot_paper}, and to actual software which is published on GitHub \cite{spiritroot_git}.

The main tasks which make up the reconstruction algorithm are:

\begin{itemize}
  \item Decoder task
  \item Pulse Shape Algorithm (PSA Task)
  \item Helix Track Finding Algorithm
  \item Track Fitting and momentum reconstruction(GENFIT package)
  \item Vertex Reconstruction (RAVE package).
\end{itemize}

\subsection{Decoder task}
The decoder task reads in the raw data output from the GET electronics and separates the data from the headers using the GETDecoder \cite{getdecoder}. A map is stored mapping each channel in the CoBo board onto the physical pad plane, the information is stored in a pad container called the STPad class. This container stores the row, layer, and the raw ADC time bucket spectrum. All of the pads combined form the event container, STRawEvent class, which represents the raw event information.

Three basic calibrations are performed at the raw data level within this task. The gain calibration calibrates the electronics will be described in Section\ref{sec:elecCalib} and the anode gain calibration is discussed in Section~\ref{sec:anodeCalib}. The gating grid noise subtraction and time bucket cut which is described in Section~\ref{sec:ggnoisesub}.
%gain calibration 
%ggnoise subtraction
%GG time bucket cut

\subsection{Pulse Shape Algorithm (PSA) Task}
\label{sec:psa}
%Yoffset match the TPC-Vertex_y with BDC_Y
%YPedestalOffset what is it for?
%cobo gain matching
An algorithm loops over the raw time bucket spectrum to find all the unique pules in each pad. It is very likely that there are several pulses in a pad coming from multiple tracks passing under the same pad which are separated only in arrival time. Figure~\ref{fig:psaTask} shows the PSA task working on one particular pad in experimental data. The shaded histogram in every sub-figure represents the ADC raw spectrum of the pad. In Fig.~\ref{fig:psa0} we see the PSA has identified the first pulse and fitted it with the standard pulse, the result of the fit is shown by the red line. The fitted red line is then subtracted from the ADC spectrum, which is shown in the shaded histogram in Fig.~\ref{fig:psa1}. Here, the second pulse is found and fitted again with the same procedure, the red line will be subtracted from the raw ADC spectrum. The black line represents the sum of all the fitted pules in each step. This continues until all pules are found or until the PSA reaches the last bin in the spectrum. Figure~\ref{fig:psa4} shows the final result where 4 fitted pulses were found. The agreement between the total summed line and the raw ADC spectrum exemplifies how successful the PSA is in this particular set of data.

%Mention about the minimum distance between hits?

The fit of each individual pulse contains the height of the pulse and the timing information of the start of the pulse, defined to be the location in time of 10\% of the rising edge. The height of the pulse, which is related to the charge induced on the pad, $Q$, and the timing information $t_o$, is saved into a new container representing a ``hit", called the STHit class. After the PSA processes each pad in the raw event, an entire hit cloud of the event is produced. The x and z-coordinates come from the center of the pad which is already stored in the map of STPad. Since the STHit class inherits from the STPad class this information is also carried over. The y-coordinate is calculated using the known drift velocity $v$ and time of the signal $t_o$ as $y = v\cdot t_0$, $t_0$. The hit class forms the fundamental structure from which all tracking is performed since it contains the x,y,z spacial coordinates, and charge information $Q$.

\begin{figure}[!htb]
     \centering
     \begin{subfigure}[b]{0.24\textwidth}
         \centering
         \includegraphics[width=\textwidth]{psa0.png}
         \caption{$\mathrm{1^{st}}$ step of the PSA}
         \label{fig:psa0}
     \end{subfigure}
     \hfill
     \centering
     \begin{subfigure}[b]{0.24\textwidth}
         \centering
         \includegraphics[width=\textwidth]{psa1.png}
         \caption{$\mathrm{2^{nd}}$ iteration}
         \label{fig:psa1}
     \end{subfigure}
     \hfill
      \centering
     \begin{subfigure}[b]{0.24\textwidth}
         \centering
         \includegraphics[width=\textwidth]{psa2.png}
         \caption{$\mathrm{3^{rd}}$ iteration}
         \label{fig:psa2}
     \end{subfigure}
     \hfill
      \centering
     \begin{subfigure}[b]{0.24\textwidth}
         \centering
         \includegraphics[width=\textwidth]{psa3.png}
         \caption{$\mathrm{4^{th}}$ iteration}
         \label{fig:psa3}
     \end{subfigure}
     \hfill
     \centering
     \begin{subfigure}[b]{.5\textwidth}
         \centering
         \includegraphics[width=\textwidth]{psa4.png}
         \caption{Final output of the PSA}
         \label{fig:psa4}
     \end{subfigure}
     \hfill

@@

\subsection{Helix Tracking}
%Cluster cut
%ellipsoid cut
\label{sec:helixtrack}

 The function of the  helix track finding algorithm is to sort hit cloud into unique sets of hits representing tracks. This is performed by a Riemann track finding algorithm where the Cartesian space is mapped onto a Riemann sphere. This is a particular transformation in which unique helices in the Cartesian space form great circles on the Riemann sphere. This is particularly useful since hits which correspond to a track form a helices the Cartesian space of the TPC coordinates. It is difficult to search for tracks in the hit cloud by fitting a general helix to different combinations of hits. It is much easier to search for collection of hits by fitting a plane to the great circles which are formed on the Riemann sphere. From here the set of hits which form a unique track are stored into a new helix track container called the STHelixTrack class.

\paragraph{Definition of clustering}
The hits within a helix track are dense and the position resolution of an individual hit is not very precise. As mentioned in Section~\ref{sec:padResponse}, the TPC can achieve sub-millimeter position resolution by utilizing the PRF. In the helix track finding task, the hits are further reduced into more precise clusters. A brief description of the method of clustering is illustrated in Figure \ref{fig:topview}. As will be seen, it is sufficient to simply cluster hits along one direction, either along the x or z-direction, which ever is most perpendicular to the local track region. For example, the three clusters at the bottom of Figure \ref{fig:topview} are clustered along the x-axis, and the upper three are along the z-axis. The bold pads highlight the hits belonging to an example cluster from each set. Since we decide to cluster only in one direction, there is a natural inflection point in a track at which the direction of clustering must switch. This depends on the crossing angle of the track $\theta$, which is defined as the angle between the tangent of the track  at that cluster and the x-axis. In this definition $\theta = \ang{90}$ when the track is going along the z-axis and $\theta = \ang{0}$ for a track going along the x-axis.  The clustering direction in the case $\ang{45} < \theta \leq \ang{90} $ is along the x-direction. For $\ang{0} < \theta \leq \ang{45}$ we cluster along the z-axis.

\begin{figure}[!htb]
\centering
\includegraphics[scale=.5]{top_view_helix_ext.pdf}
\caption{Cartoon graphic of a top down view of a fit to a track passing through several pads. The bolded pads and the charges $q_i$ represent the hits belonging to that pad and the clusters of the track representing the average position of the track. The three clusters at the bottom are clustered in the x-direction and for the upper three clustered in the z-direction. The estimate of the position of the avalanche is given by the track fit and the position from the center to each pad to the $\bar{x}$ position is given as $\lambda_i$.}
\label{fig:topview}
\end{figure}

 The position along the clustering direction, $\overline{X}$, is calculated as the charge-weighted average, $\overline{X} = \sum_i q_ix_i$, where $q_i$ and $x_i$ are the charge and position of the i-th hit in the cluster. The other direction is set to the center of the pad. For example, if we are clustering hits along the x-axis, the z-position is set to the center of the pad in the z-direction and vice versa. Switching the clustering direction gives better position resolution than if we clustered only along one direction. You can imagine if we calculated the clusters only along the x-axis for tracks with $\theta \approx 0^{\circ}$ the x-position is not well defined. The cluster position and charge is stored into a new container represented the cluster called the STHitCluster class. All the clusters which belong to a particular helix track is also stored in the corresponding STHelixTrack class.

\subsection{Correction Task}
%extend dynamic range
%PRF cut
%space charge correciton
In this task the hit distribution, within a given cluster, is checked against the measured PRF discussed in Section~\ref{sec:prf}. Figure~\ref{fig:prfpimData} shows black lines around the measured PRF in the data, which form gates defining the acceptable PRF region. If less than 50\% of the hits within a cluster lie within this gate, the cluster is thrown out from the track. A majority of the clusters contains either 2 or 3 hits in total; in the case of 2 total hits, at least one of them needs to lie in the gate, in the case of 3 total hits, at least 2 hits need to lie within the gate. Clusters which do not follow this PRF gate usually have been corrupted. Typically charges from other nearby tracks is the typical way clusters become corrupt.

%CONSIDER SHOWING SOME PROOF OF IT IN ACTION

This correction task also handles several other important corrections. The first is a novel algorithm which extends the dynamic range of the electronics and is discussed in detail in Section~\ref{sec:extendDynamicRange}. The other correction task that is handled here is the space charge correction which will be discussed in more detail in Section~\ref{sec:spacecharge}.

\subsection{Momentum and track reconstruction (GENFIT)}
After all of the corrections are applied to the clusters, the cluster's position and charge values are passed to a software package which reconstructs the momentum of the track, called GENFIT \cite{genfit}. The details of Kalman fitting algorithm used in the GENFIT reconstruction are left for the reader here \cite{genfit}. GENFIT returns a momentum and charge value for the track which is stored into a new track container called the STRecoTrack class.

The task is repeated a second time, storing a second copy of the track class but this time with the BDC vertex point added as a constraining cluster. The momentum resolution is greatly improved by adding the vertex as a constraining point; since the BDC detectors, described in Section~\ref{sec:bdc}, give such an accurate vertex location. The energy loss of the track is calculated by the truncated mean method described in Eq.~\ref{eq:truncmean}. This is the final part of the analysis which pertains to tracks, and the PID can be constructed using the energy loss, momentum, and charge information.

\subsection{Vertex tracking (RAVE)}
\label{sec:vertex}
After all tracks are reconstructed by GENFIT, the tracks are then passed to the RAVE software package which reconstructs the event vertex from a collection of tracks \cite{rave}. Only tracks which are THIS CONDITION are used in making the event vertex. The vertex location along the beam axis determines whether or not the beam collided with the target.

\section{Calibrations and Corrections}
In the following subsections we will discuss in detail the calibrations and corrections happening within the software framework that was outlined above.

\subsection{Gating grid noise subtraction}
\label{sec:ggnoisesub}
The opening of the gating grid is essentially shorting adjacent wires in the grid together, allowing them to reach equilibrium as fast as possible. Since the impedance of both sides was not entirely matched properly this caused an imbalance of charge which bounce back and forth in the grid in an under-dampened manner. This oscillating charged induced  residual induced signals early in the time bucket spectrum creating extra source of noise in the data. Figure~\ref{fig:ggNoiseSub} shows in the upper panel the gating grid noise ADC time bucket spectrum for 2000 events in a given pad. The gating grid noise is stable for a given pad and the mean value --shown as the red line-- can be calculated after averaging over several thousands events. The raw gating grid noise lasts for 100 Tbs extending into the real data with a exponentially decreasing ampli

 %The ground grid shielded some of the noise, but the path to ground was on two ends and was not sufficiency for this frequency.

@@

The gating grid noise profile is measured by recording experimental data, only turning off the anode wires. By doing so there are no signals originating from tracks in the TPC and there is only the gating grid noise, since the grid is still being triggered.  The mean value of each pad is calculated from several thousand events of the gating grid noise profile. Once the mean value response in each pad is calculated, we can subtract the mean profile from the real data and remove the noise of the gating grid.

The bottom panel of Fig.~\ref{fig:ggNoiseSub} shows the gating grid noise self correcting itself; a best case scenario. We can see the pedestal, which is around 3800 ADC in the upper pannel,  is automatically taken into account when subtracting the mean profile, and the subtracted ADC spectra is centered around 0. The residual gating grid noise that was not canceled by the subtraction remains, but is significantly reduced to a much smaller time bucket region. Most of the residual noise happens in regions where the derivative of the gating grid noise is large. In these areas the ADC value is dependent upon small jitters in the timing of the signal and as a result varies wildly. The residual noise in the subtracted spectra is removed by cutting away all the bins below 30 tb.

In the experiment the gating grid itself was quite stable in the  ${}^{132}$Sn and ${}^{124}$Sn systems and was not very stable in the ${}^{108}$Sn and ${}^{112}$Sn beams, where the gating grid broke several times. The gating grid noise was monitored and we regularly took gating grid noise profiles, especially after a gating grid board was replaced.

@@

\subsection{Cocktail calibration}
\label{sec:cocktail}

\begin{table*}\centering
\ra{1.3}
\begin{tabular}{@{}rrrrrrr@{}}\toprule
& \multicolumn{3}{c}{$100 MeV Target$}\\
\cmidrule{2-4}
Particle &\phantom{abc} & Measured & Corrected & \% Difference & \% Difference\\
\midrule
p   & 882.8 & 929.5 & 877.3   &  3.7  & -1.0  \\
d   & 817.1 & 831.15 & 797.94 &  1.7  & -2.3\\
\bottomrule
\end{tabular}
\caption{Summary of expected cocktail for the calibration run taken with the Al target. }
\label{tb:cocktail100tar}
\end{table*}

\begin{table*}\centering
\ra{1.3}
\begin{tabular}{@{}rrrrrrr@{}}\toprule
& \multicolumn{3}{c}{$100 MeV$}\\
\cmidrule{2-4}
Particle & Expected & Measured & Corrected & \% Difference & \% Difference\\
\midrule
p   & 882.8 & 903.5 & 889.0 &  2.0   & -1.6  \\
d   & 817.1 & 898.5 & 874.5 &  2.1   & -2.7\\
\bottomrule
\end{tabular}
\caption{Summary of expected cocktail from the lower beam energy.}
\label{tb:cocktail100}
\end{table*}

@@

\bottomrule
\end{tabular}
\caption{Summary of expected cocktail from the higher beam energy.}
\label{tb:cocktail300}
\end{table*}

A cocktail of several light charged particles was produced (p,d,t,${}^{3}$He,${}^{4}$He), and measured in the TPC to provide a momentum calibration of the TPC. The magnetic and slit settings of the dipoles in the BigRIPS spectrometer was set so that the momentum resolution of the beam was $\delta p/p$ < 1\%. Two magnetic rigidity settings were studied, with an empty target. A thick Aluminum target was also inserted to provide a slightly lower energy point for part of the lower rigidity setting, effectively creating three energy calibration points over several particle species. In some cases the certain particles were not able to propagate or were strongly attenuated.

Since the cocktail beam was a mono-energetic source of particles and the actual momentum resolution was less than 1\%, the observed momentum resolution can be interpreted as the intrinsic TPC momentum resolution. The momentum resolution in general depends on several factors such as the particle's angle, momentum, charge, track multiplicity, etc. This calibration beam represents and ideal situation where the track was parallel to the pad plane and only one particle was measured at one time. The average momentum resolution was measured to be $dp/p = 2\%$ over the full range of particle species for all settings.

The energy loss resolution can also be directly measured since each energy setting represents a monochromatic source of each particle species, which have a well defined energy loss distribution. The energy loss resolution was measured to be 5\% for all particles over the energy setting measured and is summarized in Table~\ref{tb:momresolution}.

Since the magnetic dipole setting of the BigRIPS spectrometer is well defined, we can calculate the expected momenta of each particle species measured. The enegy losses through various materials in the beam line were propagated using LISE++ software \cite{lise++}. The measured momenta of the calibration beam differed significantly from the expected values -- up to 5\% in the high momentum calibration -- as seen in Tables~\cref{tb:cocktail100tar,tb:cocktail100,tb:cocktail300}. This effect is attributed to inhomogenatities in the magnetic field which introduces electron drift velocities in the direction of $\vec{E}\times\vec{B}$ direction as seen from Eq.\ref{eq:elecdrift}. The $\vec{E}\times\vec{B}$ drift velocity causes the electron trajectories to shift toward the +x-axis in the TPC coordinates causing particles of positive charge (going in the -x-axis) to have a higher measured momenta than in reality.  The details of the correcting for the $\vec{E}\times\vec{B}$ effect will be discussed later in Section~\ref{sec:spacecharge} in a  general way which also includes the correction for the positive space charge. The same correction technique was applied here where the cocktail beam is the special case of zero space charge, only having $\vec{E}\times\vec{B}$ components.

The values under the corrected column in Tables~\cref{tb:cocktail100tar,tb:cocktail100,tb:cocktail300}, represent the corrected data. A significant improvement is seen in the high momentum setting going from around ~5\% disagreement to within ~1\% agreement in the corrected data. For the lower momentum settings (Tables~\cref{tb:cocktail100tar,tb:cocktail100}), protons see a slight improvement of about ~1\% where as the deuterons are over corrected in both settings. The level of agreement of the all corrected values is still well within the estimated momentum resolution of the TPC.

\begin{table*}\centering
\ra{1.3}
\begin{tabular}{@{}cc@{}}\toprule
Momentum Resolution \% & <dE/dx> Resolution \% \\
\midrule
1.6  & 4.6\\
\bottomrule
\end{tabular}
\caption{Summary of the estimated momentum and energy loss resolutions.}
\label{tb:momresolution}
\end{table*}

@@

\subsection{Electronics calibration}
\label{sec:elecCalib}

\begin{figure}[!htb]
\includegraphics[width=\linewidth]{gaincalib.png}
\caption{Repose of each channel in the electronics to various input voltages of an input signal supplied by a pulse generator.}
\label{fig:gaincalib}
\end{figure}

The electronics were calibrated by measuring the response of each channel to an input signal supplied by a pulse generator. This is a relative calibration technique with the intent to calibrate the varying gains in each channel relative to one another, and not an absolute calibration. The pulse was distributed to all the electronics channels by pulsing the ground plane for a range of input voltages. This distributed the pulse evenly across the entire pad plane. The input voltage is plotted as a function of the measured height of the pulse in the electronics channel given in units of ADC in Fig.~\ref{fig:gaincalib}, for every channel. The small variation in each channel can be seen as the wide band around each measurement point. A linear fit is performed for each channel to get the calibration line for each channel. A channel is chosen which provides a reference calibration to which all other channels are calibrated too. This is done by matching the slopes of each channel to the reference calibration slope.  The right panel shows the resulting distribution of channels after calibration.

\subsection{Anode gain calibration}
\label{sec:anodeCalib}
As discussed in Section~\ref{sec:wireplanes} the voltage of the anode sections 12 and 14 were reduced during the experiment due to high currents being observed on the wires. Out of all the runs used in the analysis in this thesis \ref{tb:runList}, the anode sections 12 and 14 were lowered to \SI{1085}{\volt} for runs 2272-2371 and set to \SI{1214}{\volt} for all the other runs. By lowering the voltage on these anode wires, the gas gain is lowered as compared with all the other anode wire plane sections which operate at \SI{1460}{\volt}. To account for the drop in gain, we increase the gain of the pads which lie above these anode wires  in the software gain of the other channels. The gain factor is found by plotting the energy loss in the high gain sections to the low gain sections.
